Fractional Types

### 2016-12-17 typespp.tex
3.7

\medskip\noindent Baez and Dolan~\cite{groupoidcard} assign to each groupoid a
\emph{cardinality} that counts the objects up to equivalences. The
groupoid on the left has six points \texttt{a}, \texttt{b}, \texttt{c},
\texttt{d}, \texttt{e}, and \texttt{f} with two groups of three
points each clustered in an equivalence class and hence the groupoid has
cardinality~2. The (2-)groupoid on the right has one point $\ast$ with
four equivalences \texttt{id}, $p$, $q$, and~$q'$ on it. The
equivalences $q$ and $q'$ are however identified by~$\alpha$ leaving
only three distinct equivalence classes and hence making the
cardinality $\frac{1}{3}$.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \otimes \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are not
categorically equivalent and hence $\simeq$ cannot be categorical equivalence.
There are however other notions of groupoid equivalence such as Morita
equivalence~\cite{looping} that identifies the two groupoids above.  Although
the notion of equivalence we use in the next section appears novel, it is
related ``in spirit'' to Morita equivalence and we find it useful to give a
small motivating example instantitating $p$ above to $\permfive$ (which has
order 3). Essentially we are verifying that the product of the groupoid
$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
with one trivial automorphism.

>>>>

\medskip\noindent Baez and Dolan~\cite{groupoidcard} assign to each groupoid a
\emph{cardinality} that counts the objects up to equivalences. The groupoid on
the left has six points \texttt{a}, \texttt{b}, \texttt{c}, \texttt{d},
\texttt{e}, and \texttt{f} with two groups of three points each clustered in an
equivalence class and hence the groupoid has cardinality~2. The (2-)groupoid on
the right has one point $\ast$ with four equivalences \texttt{id}, $p$, $q$,
and~$q'$ on it. The equivalences $q$ and $q'$ are however identified by~$\alpha$
leaving only three distinct isomorphism classes and hence making the cardinality
$\frac{1}{3}$.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite{looping}, that identify the two groupoids above.
Although the notion of equivalence we use in the next section appears novel, it
is related ``in spirit'' to Morita equivalence and we find it useful to give a
small example. We will verify that the product of the groupoid
$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
and one trivial automorphism. To avoid clutter we assume the groupoids are
strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
of 3 isomorphism classes of elements). The definition of Morita equivalence in
this case reduces to to finding a Morita morphism from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor from the
first groupoid to the next with an additional condition that it is a
\emph{pullback}. The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the pullback
condition ensures that the choice is ultimately insignificant as it leads to the
same result as every other possible functor. To summarize the equivalence
between the two groupoids requires a \emph{local} map that chooses a target and
a \emph{global} condition ensuring that the choice ultimately annhilates. The
challenge we are facing is to turn this idea into an operational semantics that
somehow combines the local and global aspects of the equivalence.
